Does an earthquake “know” how big it will be? A neural-net aided study

### 2024-04-08 article.tex
Yohai's changes

\abstract{
Earthquake occurrence is notoriously difficult to predict. Salient features of the statistical properties of the timing and location of earthquakes, i.e. spatiotemporal clustering, can be relatively well captured by point-process models. However, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts, both due to the lack of information about the fault condition and to the inherent complexity of rupture dynamics. Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events, besides the time-independent Gutenberg Richter (GR) distribution, which describes the marginal distribution over wide regions and long times. This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed. In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history. We present a neural network-based model for probabilistic forecasting of future magnitudes based on cataloged properties: hypocentral locations, occurrence times and magnitudes of past earthquakes. Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.  This demonstrates that earthquake catalogs contain information about the magnitude of future earthquakes, prior to their occurrence. We conclude by proposing methods to apply the model in characterization of the preparatory phase of earthquakes, and in operational hazard alert and earthquake forecasting systems.
}

@@

\section{Introduction} \label{sec:introduction}

Earthquake forecasting is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes \cite{bernard_earthquake_1999, geller_earthquakes_1997}. Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes. This includes various characteristics of spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}, which can be incorporated into stochastic prediction schemes \cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}. These forecasting methods focus on the rate, and sometimes the location, of earthquake occurrence [cite]. Furthermore, these spatiotemporal properties are almost always treated independently from the magnitude, an assumption known as separability \cite{schoenberg_testing_2004}. That is, while seismic history is used to inform the location and rate of future earthquakes, i.e. by assuming some phenomenological description of spatiotemporal clustering, the magnitude of future events prediction often relies on drawing values from the Gutenberg-Richter (GR) distribution or its variants \cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}. Importantly, this distribution either depends weakly or does not depend on the seismic history, implying that, conditioned on the fact that an earthquake occurs, its magnitude is taken to be independent of the rate and location of past events.

% This might be something to remove to shorten
% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice. On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff). These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics. In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).

While assuming that an earthquake's magnitude is independent of the seismic history is popular due to its simplicity and its ease of use in earthquake modeling, it is not necessarily true. Indeed, the magnitude of an earthquake is a property of the rupture process, which is governed by the frictional properties of the fault and the stress field. These in turn depend on slip history in a deterministic manner, albeit intangibly complicated. Therefore, the assumption of magnitude separability, i.e. its independent from seismic history, is not a priory justified, and its validity should be tested empirically.
Though information about an earthquake magnitude may exists before its initiation, implying dependence on seismic history and non-separability, it has yet to be shown accessible using our current experimental or analytic tools. Current research on magnitude prediction has failed to show a universal and replicable advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}, by that leaving the question of accessible information prior to an earthquake's nucleation withstanding. 
\newline
\neri{perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's}
In this paper we explore if information about an earthquake's magnitude can be obtained from regional seismic history. This is done by constructing a neural-based model that predicts the magnitude of a given earthquake, given the short and long term seismic history prior to its occurrence. Importantly, our model is not tasked with predicting the timing and location of the event, as they are explicitly provided to the model. Thus, we separate out the task of predicting the nucleation of events and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of an earthquake is extractable from cataloged properties alone.

@@

\neri{Is a literature review nescessary here or is it ingested enough in the text itself? I'm pretty satisfied with how the paragraphs up to now are written.}\newline
\neri{The entire paragraph going on about SOC, perhaps to long and should be inserted as a sentence somewhere in the literature review? can be seem as comment in source code.}
% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice. On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff). These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics. In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).

In order to tackle the question of magnitude predictability, we construct the MAGnitude prediction Neural NETwork model, MAGNNET, a neural network (NN) for producing a probability density function (PDF) of the possible magnitudes at a queried space time coordinates, conditioned on the seismic history. MAGNNET encodes the seismic history of time series from a hypocentral catalog and produces an estimate of the possible magnitudes at a queried time and place.

MAGNNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.
MAGNNET is composed of a long-short term memory (LSTM) architecture to model the seismic history up to a given time. This latent representation is then passed together with  a space-time coordinate query of a future earthquake into a fully connected neural network (FCNN) that produces a parametrized PDF of the possible magnitudes at the quarried space- time coordinates. The task described here, from encoding the seismic history to producing a PDF for a specific time-space querry is depicted in Fig. \ref{fig:intro_fig}a. A detailed description of the model's architecture is given in the methods section \neri{TODO}. We optimize MAGNNET's parameters to maximize the log likelihood (LL) of the true magnitude label, by using the loss function $\mathcal{L}$:
\begin{equation}
    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i} \left( m_i \right) \right) } \rangle\ ,
    \label{eq:loss_function}
\end{equation}
where $p_{\textbf{x}_i, t_i}$ is the probability function returned by the model for the space-time coordinates $\textbf{x}_i, t_i$ (i.e. the ``query coordinates''), $m_i$ the corresponding to the actual magnitude of the $i^{th}$ earthquake and $\langle \cdot\rangle$ stands for the empirical average over the data set. We use the a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} parameterized family of probability distributions stretched to a relevant domain. The final distribution used is defined by 5 parameters. This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value. This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.

We train MAGNNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan. While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.? cite?) and exhibit varying data quality. A separate model is trained for each region, with identical loss function and parameterization of the PDF. Sample results of known events worldwide are presented in \ref{fig:intro_fig}b. The PDF of each event is shown superimposed with the train set's Gutenberg-Richter distribution, i.e. the common vanilla benchmark.

@@

The model's, MAGNNET's, performance is evaluated over a time span that was not included in training. The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence. \neri{I think this is a good place to add that we conclude that there is some preparatory phase.}
%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation. Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.

>>>>

\abstract{
Earthquake occurrence is notoriously difficult to predict. Salient features of their spatiotemporal statistics can be relatively well captured by point-process models. However, very little is known regarding the magnitude of future events, and it is deeply debated whether it is possible to predict the magnitude of an earthquake before it starts, both due to the lack of information about fault conditions and to the inherent complexity of rupture dynamics. Consequently, even state of the art forecasting models typically assume no knowledge about the magnitude of future events besides the time-independent Gutenberg Richter distribution, which describes the marginal distribution over large regions and long times. This approach implicitly assumes that earthquake magnitudes are independent of seismic history and are identically distributed. In this work we challenge this view by showing that information about the magnitude of an upcoming earthquake can be directly extracted from the seismic history. We present a neural network-based model for probabilistic forecasting of future magnitudes based on cataloged properties: hypocentral locations, occurrence times and magnitudes of past earthquakes. Our history-dependent model outperforms stationary and quasi-stationary state of the art GR-based benchmarks, in real catalogs in Southern California, Japan and New-Zealand.  This demonstrates that earthquake catalogs contain information about the magnitude of future earthquakes, prior to their occurrence. We conclude by proposing methods to apply the model in characterization of the preparatory phase of earthquakes, and in operational hazard alert and earthquake forecasting systems.
}

@@

\section{Introduction} \label{sec:introduction}
Earthquakes are notoriously unpredictable, and forecasting seismicity is a long-standing scientific and technological challenge, often deemed unrealistic due to the inherent complexity of earthquake processes and the scarcity of near-field data~\cite{bernard_earthquake_1999, geller_earthquakes_1997}. 
Research since the late 19th century has provided much phenomenological insight about the spatiotemporal statistics of earthquakes, including various marginal distributions \yohai{[cite GR and variants]}, scaling relations \yohai{[cite]} and characteristics of both spatial and temporal clustering \cite{omori_after-shocks_1894, kagan_short-term_2004, ben-zion_localization_2020}. Clearly, these insights can be used to quantitatively inform us about future seismicity based on recent history. For example, Omori's law tells us that after big earthquakes we should expect an increase in the local seismicity rate~\cite{omori_after-shocks_1894}. Such laws have been incorporated into a variety of forecasting models, which are operationally used today\cite{ogata_statistical_1988, hardebeck_aftershock_2024, devries_deep_2018, king_static_1994}.

However, the above mentioned statistical relations only describe the rate and locations of earthquakes, and very little is known about the dependence of earthquake magnitude on seismic history.
It is in fact deeply debated whether it is possible to independent of seismic history the magnitude of an earthquake before it starts~\cite{gutenberg_frequency_1944, kagan_seismic_2002, ogata_exploring_2018}, or even during rupture~\yohai{[cite]}.
Most modeling approaches assume that event magnitudes are identically and independently drawn from a stationary (or almost stationary) distribution - the Gutenberg Richter distribution (GR), which describes the marginal magnitude distribution over large regions and long times. 
While some variations in magnitude statistics have been described \yohai{[Gulia Weimar, beta as stress sensor papers, Didier's paper]}, they all model history dependence by small and slow changes in the parameters of the GR distribution, and do not inform us about the magnitude of \textit{a specific} future event. Even state-of-the-art forecasting models typically assume no knowledge about the magnitude of future events besides the marginal GR distribution, an assumption known as separability \cite{schoenberg_testing_2004}. Indeed, Current research on magnitude prediction has failed to show a universal and reproducible advantage over the common benchmarks \cite{shcherbakov_forecasting_2019, ogata_exploring_2018, stockman_forecasting_2023,  panakkat_neural_2007}.

This modeling approach is not a naive or uninformed choice. Due to the spatial complexity of the elastic fields, faulting patterns and lithology, the nonlinearities of the rupture process and the complicated interaction between them all, it is not far-fetched to assume that determining the magnitude of an event  requires full microscopic knowledge of the system's state. 
This philosophy also inspired physical models which describe earthquake dynamics with a statistical approach \yohai{(cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff)}. These models reproduce many of the phenomenological  statisitcal relations by positing that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event is indeed impossible using only far-field measurements.

\neri{perhaps a good place to mention Giuseppe-Zhuang work, Sam's  and Kelian's} \yohai{yes}

The main goal of this paper is to ask the question directly: can we extract any information about a future earthquake's magnitude from regional seismic history? To this end, we construct a neural-based model that predicts the magnitude of a given earthquake given the short and long term seismic history prior to its occurrence. Our model is provided with seismic catalogs which include the location, time and magnitude of past events. Importantly, the model is not tasked with predicting the timing and location of the event, as they are explicitly provided. Thus, we separate out the task of modeling nucleation statistics and isolate the question of magnitude predictability. If our model performs better than a random draw from a GR distribution or its variants, as we will indeed demonstrate is the case, we assert that at least some information about the magnitude of a specific earthquake is extractable from cataloged properties alone.

@@

% This modeling approach, implicitly assuming that faults hold no information about an earthquake magnitude before its occurrence, is not a naive or uninformed choice. On the contrary, it is supported by a broad range of physical models which describe earthquake statistics as a critical phenomenon (cite OFC, Bak's papers, Sornette 91, Sornette 2006, Wyart's new stuff). These models posit that faults evolve (``self-organize'') towards a critical state, where events emerge stochastically and their magnitude follows a power-law distribution which is scale-free and self-similar, akin to physical systems in the vicinity of a phase transition. Under this paradigm, determining the magnitude of an event requires a full microscopic knowledge of the system's state, due to the chaotic nature of rupture dynamics. In accord, point-process models predict (stochastically) the location and timing of earthquakes, but their magnitude is drawn from a constant or slowly evolving distribution (cite ETAS, other magnitude models).

\section{Methodology}
We construct the MAGnitude Neural EsTimation model, MAGNNET, a generative neural network (NN) that receives as input a hypocentral catalog of past regional seismicity, and the time and location of a future event (a ``query''). The network produces a probability density function (PDF) estimating its magnitude.

%MAGNNET's result will be quantitatively compared to other known benchmarks to estimate the amount of information gained by encoding the seismic history.

MAGNNET is composed of two main components. First, the catalog up to time $t$ is passed through a long-short term memory (LSTM) unit, which produces a latent representation of the seismic history.
Second, this representation combined with a space-time query specifying the coordinates of a future earthquake and passed into a fully connected neural network (FCNN) that produces a parametrized PDF of the event's magnitude, see Fig. \ref{fig:intro_fig}a. A detailed description of the model's architecture is given in the methods section \neri{TODO}. We optimize MAGNNET's parameters to maximize the log likelihood (LL) of the observed magnitudes,
\begin{equation}
    \mathcal{L} = -\langle \log{ \left( p_{\textbf{x}_i, t_i(h_t)} \left( m_i \right) \right) } \rangle\ ,
    \label{eq:loss_function}
\end{equation}
where $\bf{x}_i$, $t_i$, $m_i$ re the location, time and magnitude of the $i$-th event, $p_{\textbf{x}_i, t_i}(h_t)$ is the probability function returned by the model,and $h_t$ is the seismic history up to time $t$. $\langle \cdot\rangle$ stands for the empirical average over all earthquakes in the data set. Importantly, during training we only supply the model with queries about spacetime coordinates where earthquakes indeed occured. \yohai{we should state that we only consider events about $m_c$}

Technically, the PDF is a mixture of two Kumaraswamy distributions \cite{kumaraswamy_generalized_1980} stretched to a relevant domain, which is a parameteric distribution with 5 parameters, that the network outputs. This family smoothly can interpolate between a decaying distribution (resembling the GR distribution), and localized distributions whose mass is concentrated around a specific value. This allows the model to output both an ``ignorant'' prediction, essentially resembling the GR distribution, and more confident predictions localized around a given magnitude.

We deployed MAGNNET on three distinct earthquake catalogs to assess the performance across diverse seismogenic regions: the Hauksson Catalog [cite] for Southern California, GeoNet [cite] for New Zealand, and the JMA catalog [cite] for Japan. While all three catalogs encompass highly active seismic zones, they are compiled using various measurement methodologies (e.g.? cite?) and exhibit varying data quality. A separate model is trained for each region, with identical loss function and parameterization of the PDF. The resulting PDF is presented in \ref{fig:intro_fig}b for a few examples of major events, superimposed on the stationary GR distribution (fitted on the train set), which is the naive benchmark.

@@

The model's, MAGNNET's, performance is evaluated over a time span that was not included in training. The presented metrics demonstrate that our model consistently and significantly outperforms all benchmarks across all test regions, indicating an information gain in forecasting earthquake magnitudes prior to their occurrence. \neri{I think this is a good place to add that we conclude that there is some preparatory phase.} \yohai{This is not a conclusion that you can draw from our result. We can only say that there is some information about the magnitude of an event in the seismic history.} \neri{I think we can say that the information gain is indicative of a preparatory phase.} \yohai{I think we should be careful with this.} \neri{I think we should discuss this.} \yohai{This is quite amazing, I only wrote the first sentence (``This is not a conclusion that you can draw from our result'') and copilot autocompleted the rest of the dialog, correctly using the \textbackslash neri\{...\} and \textbackslash yohai\{\} macros }
%This finding directly suggests that the seismic system undergoes preparatory processes tailored to the impending event's magnitude before its initiation. Examining these results across multiple catalogs and regions highlights the robustness of our methodology and the generalizability of our conclusion, paving the way for further exploration of earthquake predictability.
